Constructing a universe for the setoid model

### 2021-01-21 paper.tex
Typo.

We conclude the paper with a discussion of further work in
Section~\ref{further-work}.

@@

The setoid model was first described in \cite{hofmann} in order to add
extensionality principles to Type Theory such as function extensionality and
propositional extensionality. A strict variant of the setoid model was given in
\cite{setoid99} using a definitionally proof-irrelevant universe of
propositions. Recently, support for such a universe was added to the
proof-assistants Agda and Coq \cite{agda-prop}, allowing a full formalization of
Altenkirch's setoid model. Setoid Type Theory (SeTT) is a recently developed
formal system derived from this model construction \cite{mpc19}. Observational
Type Theory (OTT) \cite{alti:ott-conf} is a syntax for the setoid
model differing from SeTT in the use of a different notion of heterogeneous
equality. Moreover, the consistency proof for OTT relies on Extensional Type
Theory, whereas for SeTT it is obtained via a syntactic translation.
%
XTT \cite{xtt} is a cubical variant of OTT where the equality type is defined
using an interval pretype
\\footnote{To quote one of the referees: \emph{the fact that the
    interval is a pretype is but the easiest part of the story.}}
. XTT's universes support universe induction, whereas
it is left open whether the construction presented here supports this principle.
%
Palmgren and Wilander \cite{2014arXiv1408.1364P} construct a setoid
universe using a translation into constructive set theory. Palmgren
\cite{2019arXiv190901414P} constructs an encoding of ETT in ITT
through Aczel's encoding of set theory in type theory
\cite{ACZEL197855}. He uses type theory as a language for his
formalisation but his construction is set-theoretic in nature. Setoids
are utilized to encode sets as arbitrarily branching well-founded
trees quotiented by bisimulation. His notion of family of setoids does
not use strict propositions and it has a weaker form of proof
irrelevance which seems to be not enough to obtain a model of SeTT.

@@

An exceptional aspect of the metatheory used in this paper is the presence of a
proof-irrelevant identity type with a strong transport rule allowing to
eliminate into arbitrary types.
%
In \cite{abel:nbe09}, Abel gives a proof of normalization for the
Logical Framework extended with a similar proof-irrelevant equality type.
%
Abel and Coquand show in \cite{abel2019failure} that the combination of
impredicativity with a strong transport rule results in terms that fail to
normalize.

@@

We will follow the method outlined in Section~\ref{ir-universes}. In
addition to $\inU$ for defining $\U$, we also introduce a family
$\inUU$ of binary relations between types in the universe, from which we
then define $\_\sim_{\setoidU}\_$.
\vspace{-0.2em}
\begin{align*}
  & \textsf{data}\ \inU : \mType \to \mType_1 \\
  & \qquad \textsf{bool} : \inU\ \mathbbm{2} \\
  & \qquad \pi : \forall\{A\;A_\sim\;B\;B_\sim\}(a : \inU\;A)
  (b : (x : A) \to \inU\;(B\;x))\\
  & \qquad\ \to \inUU\;a\;a\;A_\sim \to (\forall\{x_0\;x_1\}(x_{01} : A_\sim\;x_0\;x_1)
  \to \inUU\;(b\;x_0)\;(b\;x_1)\;(B_\sim\;x_{01})) \\
  & \qquad\ \to \inU\;
  (\Sigma\;(f : (x : A) \to B\;x) \\
  &\qquad\qquad\qquad\quad\, ((x_0\;x_1 : A)(x_{01} : A_\sim\;x_0\;x_1) \to B_\sim\;x_{01}\;(f\;x_0)\;(f\;x_1))) \\
  %%%%%%
  & \textsf{data}\ \inUU : \{A\ A' : \mType\} \to \inU\ A \to \inU\ A' \to (A \to A' \to \mProp) \to \mType_1 \\
  & \qquad \mathsf{bool}_\sim : \inUU\;\textsf{bool}\;\textsf{bool}\;
  (\lambda x_0\;x_1\; .\; x_0 \stackrel{?}{=}_{\mathbbm{2}} x_1) \\
  & \qquad \pi_\sim :
  \forall\{A_0\;A_1\;A_{0\sim}\;A_{1\sim}\;A_{01\sim}\;B_0\;B_1\;B_{0\sim}\;B_{1\sim}\;B_{01\sim}\}
  \{ a_0 : \inU\;A_0 \}\{ a_1 : \inU\;A_1 \}\\
  & \qquad\quad\quad\ \; \{b_0 : (x_0 : A_0) \to \inU\;(B_0\;x_0)\}
  \{b_1 : (x_1 : A_1) \to \inU\;(B_1\;x_1)\}\\
  & \qquad\quad\quad\ \; \{a_{0\sim} : \inUU\;a_0\;a_0\;A_{0\sim}\}
  \{a_{1\sim} : \inUU\;a_1\;a_1\;A_{1\sim}\} \\
  & \qquad\quad\quad\ \; \{ b_{0\sim} : \forall\{x_0\;x_1\}(x_{01} : A_{0\sim}\;x_0\;x_1)
    \to \inUU\;(b_0\;x_0)\;(b_0\;x_1)\;(B_{0\sim}\;x_{01}) \}\\
  & \qquad\quad\quad\ \; \{ b_{1\sim} : \forall\{x_0\;x_1\}(x_{01} : A_{1\sim}\;x_0\;x_1)
    \to \inUU\;(b_1\;x_0)\;(b_1\;x_1)\;(B_{1\sim}\;x_{01}) \}\\
  & \qquad\quad \to \inUU\;a_0\;a_1\;A_{01\sim} \\
    & \qquad\quad \to (\forall\{x_0\;x_1\}(x_{01} : A_{01\sim}\;x_0\;x_1)
      \to \inUU\;(b_0\;x_0)\;(b_1\;x_1)\;(B_{01\sim}\;x_{01})) \\
  & \qquad\quad \to \inUU\;(\pi\;a_0\;a_{0\sim}\;b_0\;b_{0\sim})\;
      (\pi\;a_1\;a_{1\sim}\;b_1\;b_{1\sim}) \\
  & \qquad\qquad\qquad\quad\,(\lambda f_0\;f_1\;.\;\forall(x_0\;x_1)\to A_{01\sim}\;x_0\;x_1
        \to B_{01\sim}\;x_{01}\;(\pi_1\;f_0\;x_0)\;(\pi_1\;f_1\;x_1))
\end{align*}

@@

Unfortunately this definition only manages to capture the idea, but does not
actually typecheck. In fact, $\_\sim_{\setoidU}\_$ should be $\mProp_1$-valued
relation, so $A \sim_{\setoidU} B$ should be a proposition. However, the
$\Sigma$ type shown above clearly isn't, since it quantifies over a type of
relations, which is not a proposition.
%
One possible solution is actually quite simple, and it just involves truncating
the $\Sigma$ type above to force it to be in $\mProp_1$.
\vspace{-0.2em}
\begin{align*}
  & \_\sim_{\setoidU}\_ : \setoidU \to \setoidU \to \mProp_1 \\
  & (A , a) \sim_{\setoidU} (B , b) =
   \ \parallel \Sigma\ (R : A \to B \to \mProp)\ (\inUU\ a\ b\ R) \parallel
\end{align*}

>>>>

% We conclude the paper with a discussion of further work in
% Section~\ref{further-work}.

@@

The setoid model was first described in \cite{hofmann} in order to add
extensionality principles to Type Theory such as function extensionality and
propositional extensionality. A strict variant of the setoid model was given in
\cite{setoid99} using a definitionally proof-irrelevant universe of
propositions. Recently, support for such a universe was added to the
proof-assistants Agda and Coq \cite{agda-prop}, allowing a full formalization of
Altenkirch's setoid model. Setoid Type Theory (SeTT) is a recently developed
formal system derived from this model construction \cite{mpc19}. Observational
Type Theory (OTT) \cite{alti:ott-conf} is a syntax for the setoid
model differing from SeTT in the use of a different notion of heterogeneous
equality. Moreover, the consistency proof for OTT relies on Extensional Type
Theory, whereas for SeTT it is obtained via a syntactic translation.
%
XTT \cite{xtt} is a cubical variant of OTT where the equality type is defined
using an interval pretype
\footnote{To quote one of the referees: \emph{the fact that the
    interval is a pretype is but the easiest part of the story.}}
. XTT's universes support universe induction, whereas
it is left open whether the construction presented here supports this principle.
%
Palmgren and Wilander \cite{2014arXiv1408.1364P} construct a setoid
universe using a translation into constructive set theory. Palmgren
\cite{2019arXiv190901414P} constructs an encoding of ETT in ITT
through Aczel's encoding of set theory in type theory
\cite{ACZEL197855}. He uses type theory as a language for his
formalisation but his construction is set-theoretic in nature. Setoids
are utilized to encode sets as arbitrarily branching well-founded
trees quotiented by bisimulation. His notion of family of setoids does
not use strict propositions and it has a weaker form of proof
irrelevance which seems to be not enough to obtain a model of SeTT.

@@

An exceptional aspect of the metatheory used in this paper is the presence of a
proof-irrelevant identity type with a strong transport rule allowing to
eliminate into arbitrary types.
%
In \cite{abel:nbe09}, Abel gives a proof of normalization for the
Logical Framework extended with a similar proof-irrelevant equality type.
%
Abel and Coquand show in \cite{abel2019failure} that the combination of
impredicativity with a strong transport rule results in terms that fail to
normalize but this is irrelevant in our setting.

@@

We will follow the method outlined in Section~\ref{ir-universes}. In
addition to $\inU$ for defining $\U$, we also introduce a family
$\inUU$ of binary relations between types in the universe, from which we
then define $\_\sim_{\setoidU}\_$.
{\small\vspace{-0.2em}
\begin{align*}
  & \textsf{data}\ \inU : \mType \to \mType_1 \\
  & \qquad \textsf{bool} : \inU\ \mathbbm{2} \\
  & \qquad \pi : \forall\{A\;A_\sim\;B\;B_\sim\}(a : \inU\;A)
  (b : (x : A) \to \inU\;(B\;x))\\
  & \qquad\ \to \inUU\;a\;a\;A_\sim \to (\forall\{x_0\;x_1\}(x_{01} : A_\sim\;x_0\;x_1)
  \to \inUU\;(b\;x_0)\;(b\;x_1)\;(B_\sim\;x_{01})) \\
  & \qquad\ \to \inU\;
  (\Sigma\;(f : (x : A) \to B\;x) \\
  &\qquad\qquad\qquad\quad\, ((x_0\;x_1 : A)(x_{01} : A_\sim\;x_0\;x_1) \to B_\sim\;x_{01}\;(f\;x_0)\;(f\;x_1))) \\
  %%%%%%
  & \textsf{data}\ \inUU : \{A\ A' : \mType\} \to \inU\ A \to \inU\ A' \to (A \to A' \to \mProp) \to \mType_1 \\
  & \qquad \mathsf{bool}_\sim : \inUU\;\textsf{bool}\;\textsf{bool}\;
  (\lambda x_0\;x_1\; .\; x_0 \stackrel{?}{=}_{\mathbbm{2}} x_1) \\
  & \qquad \pi_\sim :
  \forall\{A_0\;A_1\;A_{0\sim}\;A_{1\sim}\;A_{01\sim}\;B_0\;B_1\;B_{0\sim}\;B_{1\sim}\;B_{01\sim}\}
  \{ a_0 : \inU\;A_0 \}\{ a_1 : \inU\;A_1 \}\\
  & \qquad\quad\quad\ \; \{b_0 : (x_0 : A_0) \to \inU\;(B_0\;x_0)\}
  \{b_1 : (x_1 : A_1) \to \inU\;(B_1\;x_1)\}\\
  & \qquad\quad\quad\ \; \{a_{0\sim} : \inUU\;a_0\;a_0\;A_{0\sim}\}
  \{a_{1\sim} : \inUU\;a_1\;a_1\;A_{1\sim}\} \\
  & \qquad\quad\quad\ \; \{ b_{0\sim} : \forall\{x_0\;x_1\}(x_{01} : A_{0\sim}\;x_0\;x_1)
    \to \inUU\;(b_0\;x_0)\;(b_0\;x_1)\;(B_{0\sim}\;x_{01}) \}\\
  & \qquad\quad\quad\ \; \{ b_{1\sim} : \forall\{x_0\;x_1\}(x_{01} : A_{1\sim}\;x_0\;x_1)
    \to \inUU\;(b_1\;x_0)\;(b_1\;x_1)\;(B_{1\sim}\;x_{01}) \}\\
  & \qquad\quad \to \inUU\;a_0\;a_1\;A_{01\sim} \\
    & \qquad\quad \to (\forall\{x_0\;x_1\}(x_{01} : A_{01\sim}\;x_0\;x_1)
      \to \inUU\;(b_0\;x_0)\;(b_1\;x_1)\;(B_{01\sim}\;x_{01})) \\
  & \qquad\quad \to \inUU\;(\pi\;a_0\;a_{0\sim}\;b_0\;b_{0\sim})\;
      (\pi\;a_1\;a_{1\sim}\;b_1\;b_{1\sim}) \\
  & \qquad\qquad\qquad\quad\,(\lambda f_0\;f_1\;.\;\forall(x_0\;x_1)\to A_{01\sim}\;x_0\;x_1
        \to B_{01\sim}\;x_{01}\;(\pi_1\;f_0\;x_0)\;(\pi_1\;f_1\;x_1))
\end{align*}}

@@

Unfortunately this definition only manages to capture the idea, but does not
actually typecheck. In fact, $\_\sim_{\setoidU}\_$ should be a $\mProp_1$-valued
relation, so $A \sim_{\setoidU} B$ should be a proposition. However, the
$\Sigma$ type shown above clearly isn't, since it quantifies over a type of
relations, which is not a proposition.
%
One possible solution is actually quite simple, and it just involves truncating
the $\Sigma$ type above to force it to be in $\mProp_1$.
\vspace{-0.2em}
\begin{align*}
  & \_\sim_{\setoidU}\_ : \setoidU \to \setoidU \to \mProp_1 \\
  & (A , a) \sim_{\setoidU} (B , b) =
   \ \parallel \Sigma\ (R : A \to B \to \mProp)\ (\inUU\ a\ b\ R) \parallel
\end{align*}
